Empowering the use of community driven preprints in ecology and evolution

### 2024-04-08 ms/ms.qmd
Some writing

# Load packages
#install.packages("pacman")
pacman::p_load(tidyverse, here, patchwork, readxl, MASS, googlesheets4, janitor, maps, forcats,RColorBrewer, flextable, ggrepel, ggimage, gt, performance, genderdata, gender, ggthemes, mapproj, DemografixeR, countrycode, see, glmmTMB, cowplot)

@@

@@

After removing `r dim(dupl)[1]` duplicate titles – suggesting a few authors created multiple submissons for the same preprint rather than updating an existing submission – we had data for a total of `r pr` articles with ~55-60 preprints published each month in the last two years ([@fig-summary]A). Articles covered all major taxomic groups, with the most common being plants (`r sum_taxa$perc[1]`%), vertebrates (`r sum_taxa$perc[6]`%), and invertebrates (`r sum_taxa$perc[5]`%) ([@fig-summary]). A total of `r length(unique(countries$country))-1` countries have used *EcoEvoRxiv* as a preprint server. Unsurprisingly, countries in North America and Europe use articles more exstensively. Countries in Africa, Central America and parts of Asia are still slow to use *EcoEvoRxiv* ([@fig-summary]). Overall, 90% of the articles come from just `r nrow(most_use)-1` countries ([@fig-summary]).

Given that *EcoEvoRxiv* is a community driven preprint server it has managed to diversify the types of articles accepted, including reports and books that are not easily accessible but may contain important data of relevance to research syntheses. Accepting a greater diversity of artciles types allows *EcoEvoRxiv* to begin dealing with the challenges of including 'grey literature'. As a result, articles on *EcoEvoRxiv* were diverse in nature, with research articles (`r sum_types[sum_types$type == "research article",]$perc`%) being the most common, followed by reviews and meta-analyses (`r sum_types[sum_types$type == "reviews and meta-analyses",]$perc`%) and opinion papers (`r sum_types[sum_types$type == "opinion",]$perc`%) ([@fig-summary]). Reports, parcticularly from government and industry research areas, are not yet common on *EcoEvoRxiv*, and we would encourage authors to consider posting these types of articles to ensure they are openly accessible. *EcoEvoRxiv* has already formed fruitful connections with the International Union for Conservation of Nature (IUCN) to be a place where their reports can be lodged to make them more easily accessible to the broader research community. For example, IUCN Red Ecosystem Reports will be lodged on *EcoEvoRxiv* .

@@

Preprints can be a useful way to ensure that research remains open and accessible to the research community as the paper is available online no matter what the published status of the preprint ends up being. Most of the published preprints ended up being open access(@fig-pubsummary A), however, the status of many (i.e., `r sum_is_oa$n[3]` or `r sum_is_oa$percent[3]*100`%) were unknown, and `r sum_is_oa$n[1]` or `r sum_is_oa$percent[1]*100`% were not published in open access reserach journals. The open access status of the papers that were published in open access journals also varied widely (@fig-pubsummary B).

Data and code sharing is a key component of open science. Despite this, we found that `r preprint_data[1,2]` (`r preprint_data[1,"percent"]*100`%) of relevant preprints on *EcoEvoRxiv* (i.e., only where data was generated) did not share data, while `r preprint_code[1,2]` (`r preprint_code[1,"percent"]*100`%) did not share code.

@@

>>>>

# Load packages
#install.packages("pacman")
pacman::p_load(tidyverse, here, patchwork, readxl, MASS, googlesheets4, janitor, maps, forcats,RColorBrewer, flextable, ggrepel, ggimage, gt, performance, genderdata, gender, ggthemes, mapproj, DemografixeR, countrycode, see, glmmTMB, cowplot, viridis)

@@

# *EcoEvoRxiv* as a community driven preprint server three years on: evaluating novel initiatives to improve open science and break down barriers to publishing

@@

After removing `r dim(dupl)[1]` duplicate titles – suggesting a few authors created multiple submissons for the same preprint rather than updating an existing submission – we had data for a total of `r pr` articles with ~55-60 preprints published each month in the last two years ([@fig-summary]A). Articles covered all major taxomic groups, with the most common being plants (`r sum_taxa$perc[1]`%), vertebrates (`r sum_taxa$perc[6]`%), and invertebrates (`r sum_taxa$perc[5]`%) ([@fig-summary]). A total of `r length(unique(countries$country))-1` countries have used *EcoEvoRxiv* as a preprint server. Unsurprisingly, countries in North America and Europe preprint more exstensively. Countries in Africa, Central America and parts of Asia are still slow to use *EcoEvoRxiv* ([@fig-summary]). Overall, 90% of the articles come from just `r nrow(most_use)-1` countries ([@fig-summary]).

Given that *EcoEvoRxiv* is a community driven preprint server it has managed to diversify the types of articles accepted. It accepts reports and books that are not easily accessible but may contain important data of relevance to research syntheses. Accepting a greater diversity of article types allows *EcoEvoRxiv* to begin dealing with the challenges of finding 'grey literature'. As a result, articles on *EcoEvoRxiv* are diverse in nature, with the typical research articles (`r sum_types[sum_types$type == "research article",]$perc`%) being the most common, followed by reviews and meta-analyses (`r sum_types[sum_types$type == "reviews and meta-analyses",]$perc`%) and opinion papers (`r sum_types[sum_types$type == "opinion",]$perc`%) ([@fig-summary]).

Reports, parcticularly from government and industry research areas, are not yet common on *EcoEvoRxiv*. We would encourage authors to consider posting these types of articles to ensure they are openly accessible. *EcoEvoRxiv* has already formed fruitful partnerships with the International Union for Conservation of Nature (IUCN) to be a place where their reports can be lodged to make them more easily accessible to the broader research community. For example, IUCN Red Ecosystem Reports will be posted on *EcoEvoRxiv* and *EcoEvoRxiv* has worked closely with the IUCN to ensure these reports meet the IUCN requirements.

Multilingual initiatives have also been slow to take off on *EcoEvoRxiv*, with only a few Spanish preprints posted to date, and none in our other accepted langauges (e.g., French, and Portuguese). Part of the challenge in getting authors to submit multilingual preprints is the lack of awareness of *EcoEvoRxiv* in non-English speaking countries; however, cultural differences in the perception of preprints and a strong reliance on traditional publishing models, which typically favour publishing in English, may also be contributing to the slow uptake.

# Generational gaps of preprinting in ecology and evolution

# Following the journey of a preprint on *EcoEvoRxiv*: from submission to publication, and beyond

@@

Preprints can be a useful way to ensure that research remains open and accessible to the research community as the paper is available online no matter what the published status of the preprint ends up being. Most of the published preprints ended up being open access(@fig-pubsummary A), however, the status of many (i.e., `r sum_is_oa$n[3]` or `r sum_is_oa$percent[3]*100`%) were unknown, and `r sum_is_oa$n[1]` or `r sum_is_oa$percent[1]*100`% were not published in open access reserach journals. The open access status of the papers that were published in open access journals also varied widely (@fig-pubsummary B). Data and code sharing is a key component of open science. Despite this, we found that `r preprint_data[1,2]` (`r preprint_data[1,"percent"]*100`%) of relevant preprints on *EcoEvoRxiv* (i.e., only where data was generated) did not share data, while `r preprint_code[1,2]` (`r preprint_code[1,"percent"]*100`%) did not share code.

@@

# Paving *our* own future of open, transparent and community driven science

Despite the early successes of new initiatives taken by *EcoEvoRxiv* it is still clear that much work needs to be done to improve authors' understanding of preprints (and postprints). Our community needs to work harder to help educate our colleagues in this space. We view this perspective piece as one small step in achieving that goal. Our data driven approach has allowed us to better understand some of the key elements of uncertainty surrounding preprints which provides us with opportunties to clarify our perspective and recommendations. First, authors need not be afraid of sharing data and code early in the publication process. Sharing data and code early can help improve the quality of research, establish precedence, and improve the transparency and computational reproducibility of scientific findings. If authors are worried about data being used unintentially, clear information surrounding its re-use and withholding some data or meta-data needed to understand the dataset may be options. Second, authors need to be aware of the benefits of posting preprints and postprints. Posting preprints can help establish precedence of research findings, ensure that research is open access, and help authors receive feedback on their research before it is published. Third, authors need to be aware of the benefits of using community driven preprint servers. Community driven preprint servers can help ensure that research is open access, that research is reviewed by experts in the field, and that research is disseminated quickly.
